---
title: "p8105_sd3731_hw2"
author: "Susie Dong"
date: "2023-10-04"
output: github_document
---

```{r setup}
library(tidyverse)
library(readxl)
```

## Problem 1

1.Tidy 'pols-month.csv'

```{r}
pols <- 
  # load data
  read_csv("./hw2_data/P1_fivethirtyeight_datasets/pols-month.csv") |> 
  # clean column name
  janitor::clean_names() |> 
  # separate date into 'year', 'month', 'day'
  separate(mon, c("year", "month","day"), convert = TRUE, sep = '-') |> 
  # change month number to month name
  mutate(month = month(month, label = TRUE), 
         # create 'president' variable according to value of 'prez_gop' variable
         president = recode(
           prez_gop, 
           "0" = "dem", 
           "1" = "gop",
           "2" = "gop"
         )) |>
  # remove some columns
  select(-c(day, prez_dem, prez_gop)) 
```

2.Tidy 'snp.csv'

```{r}
snp <- 
  # load data
  read_csv("./hw2_data/P1_fivethirtyeight_datasets/snp.csv", 
           col_types = cols(date = col_date(format = "%m/%d/%y"))) |>
  # clean column name
  janitor::clean_names() |>  
  # separate date into 'year', 'month', 'day'
  separate(date, c("year", "month","day"), convert = TRUE) |>
  mutate(
    # change month number to month name
    month = month(month, label = TRUE),
    # change year format
    year = if_else(year > 2023, year - 100, year)) |> 
  # arrange columns
  arrange(year, month) |> 
  # remove some columns
  select(-day)
```

3.Tidy 'unemployment.csv'

```{r}
unemploy <- 
  # load data
  read.csv("./hw2_data/P1_fivethirtyeight_datasets/unemployment.csv") |>
  # clean column name
  janitor::clean_names()  |>
  # switch "wide" to "long" format
  pivot_longer(-year, names_to = "month", values_to = "unemploy_rate") |> 
  # capitalize first letter to ensure same values in key variable
  mutate(month = str_to_title(month))  
```

### data merge

```{r}
#Join datasets (Merge)
res <- left_join(pols, snp, by = join_by(month == month, year == year))
res <- left_join(res, unemploy, by = join_by(month == month, year == year))
```

## Problem 2

Read and clean the Mr. Trash Wheel sheet:

*specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
*use reasonable variable names
*omit rows that do not include dumpster-specific data

1.Tidy "Mr.Trash Wheel" sheet

```{r}
trash1 <- 

  readxl::read_excel("./hw2_data/P2_Trash Wheel Collection Data.xlsx", sheet = 1, skip = 1) |> 

  janitor::clean_names() |>

  janitor::remove_empty(which = "cols") |>
  
  filter(!is.na(dumpster)) |>

  mutate(homes_powered = 500 * weight_tons + 30,

         year = as.character(year))
```

There are `r nrow(trash1)` observations and `r ncol(trash1)` variables in the datasets.

2.Tidy "Professor Trash Wheel" sheet

```{r}
trash2 <- 
  
  readxl::read_excel("./hw2_data/P2_Trash Wheel Collection Data.xlsx", sheet = 2, skip = 1) |> 

  janitor::clean_names() |>
  
  janitor::remove_empty(which = "cols") |>
  
  filter(!is.na(dumpster)) |>

  mutate(homes_powered = 500 * weight_tons + 30,
       
         year = as.character(year))
```

There are `r nrow(trash2)` observations and `r ncol(trash2)` variables in the datasets.

3.Tidy "Gwynnda Trash Wheel" sheet

```{r}
trash3 <- 
  
  readxl::read_excel("./hw2_data/P2_Trash Wheel Collection Data.xlsx", sheet = 3, skip = 1) |> 
  
  janitor::clean_names() |>
  
  janitor::remove_empty(which = "cols") |>
  
  filter(!is.na(dumpster)) |>
  
  mutate(homes_powered = 500 * weight_tons + 30,
         
         year = as.character(year))
```

There are `r nrow(trash3)` observations and `r ncol(trash3)` variables in the datasets.

### Data Combination

```{r}
total_trash <- bind_rows(
  list("Mr. Trash Wheel" = trash1, 
       "Professor Trash Wheel" = trash2, 
       "Gwynnda Trash Wheel" = trash3), .id = "source")
```

There are `r nrow(total_trash)` observations and `r ncol(total_trash)` variables in the datasets. 

### questions:  

The total weight of trash collected by Professor Trash Wheel equals to `r sum(pull(trash2, weight_tons))`. 

The total number of cigarette butts collected by Gwynnda in July of 2021
equals to `r trash3 |> filter(month == "July" & year == 2021) |> pull(cigarette_butts) |> sum()`.

## Problem 3


